---
title: "Xenium Seurat Object Navigation"
format: 
  gfm:
    highlight-style: tango
    toc: true
    wrap: auto
    fig-width: 8
    fig-height: 5
execute: 
  warning: false
  
---

This document is meant to be a reference for working with the Xenium Seurat object output from the nf_xpatial pipeline. It contains information about what is included in the object as well as a demo for how to select various parameters for visualizing the data.

## Seurat Object Structure

Each normalization method (e.g. log-normalization, area normalization) is stored as a separate Seurat object to reduce the (already large) size of the object. Clustering metadata as well as dimensionality reductions are specific for each normalization method, so they are stored within the respective Seurat object.

For basic information about the structure of a Seurat object, please see the [Seurat wiki](https://github.com/satijalab/seurat/wiki/seurat).

### Metadata

The Seurat object contains all standard information expected in a Seurat object, as well as a number of metadata columns specific to the Xenium data.

Clustering parameters follow a pattern based the following general format: `clust_[method]_[clustering parameters]`

Current methods include:

-   `SEU`: Seurat clustering
-   `BSKY`: BANKSY clustering
-   `BSKYSEU`: BANSKY's Seurat wrapper

Clustering parameter options include:

-   `d`: Number of PCA dimensions kept for downstream processes (e.g. Harmony, UMAP, etc.)
-   `r`: Clustering resolution
-   `l`: Banksy's `lambda` parameter. It controls how much weight to give to a spatial neighborhood's transcriptome vs. an individual cell's transcriptome.
-   `k`: Banksy's `k_geom` parameter. It controls the number of cells used to create the spatial neighborhood for each cell.

Examples:

-   Seurat clustering: `clust_SEU_d30_r0.7` represesnts Seurat clustering (after Harmony integration) with 30 dimensions and a resolution of 0.7.
-   Banksy clustering: `clust_BSKY_l0.9_k15_d30_r1.0` represents Banksy clustering with a lambda of 0.9, k_geom (spatial neighborhood size) of 15, 30 dimensions of PCA / Harmony integration utilized, and a clustering resolution of 1.0.

### Reductions

Currently, each compiled object already contains all calculated dimensionality reductions for each clustering parameter (within that normalization method). This does make a large object, but allows for faster visualization and exploration of the data without needing to reload the reductions from disk. We do [provide a helper function available here](https://github.com/U-BDS/nf_xpatial/blob/main/assets/filter_xenium_obj.R) which allows a user to filter down the object to retain only selected reductions/clustering parameters. An example of its usage is shown later in this demo.

The reductions also follow a similar naming pattern to the clustering metadata, currently with the following general format: `[method]_[reduction]_[clustering parameters]`

Current methods include:

-   `Seurat`: Seurat clustering
-   `BANKSY`: BANKSY clustering
-   `BANKSYSeurat`: BANSKY's Seurat wrapper

Reductions included are:

-   `pca`: Principal Component Analysis
    -   Note: PCA reduction names for Seurat-based methods (BANKSY and BANKSY's Seurat wrapper) contain the normalization method (e.g. `pcalognorm`, `pcaareanorm`)
-   `harmony`: Harmony integration (batch corrected PCA)
-   `umap`: Uniform Manifold Approximation and Projection

Clustering parameter options match those seen in the metadata above, with the exception that reductions are calculated prior to clustering, so there are no resolution (`r`) parameters in their names.

## Visualization

Here we will demo a basic visualization of a UMAP and spatial plot for the current log-normalized Seurat object.

```{r, message=FALSE}
library(Seurat)
library(ggplot2)
library(Polychrome)
library(patchwork)

seurat_object <- readRDS("./compiled_log_norm_all_clusters.rds")
```

### UMAP

```{r}
# Default Seurat colors are often difficult to distinguish between for higher number of clusters, so we recommend Polychrome or related packages to generate more flexible color palletes.
clust_colors <- Polychrome::createPalette(
  N = length(unique(seurat_object$clust_SEU_d30_r0.5)), 
  seedcolors = c("#FF0000","#00FF00","#0000FF")
  ) |> as.character()

DimPlot(seurat_object, 
        reduction = "Seurat_umap_d30",
        group.by = "clust_SEU_d30_r0.5") + 
  scale_color_manual(values = clust_colors)
```

### Spatial Plot

```{r, }
# By default, Image coordinates are stored both flipped and rotated to their original positions on the Xenium slides, so we include some additional steps here to rotate things to their proper orientation for all FOVs.

plot_list <- ImageDimPlot(seurat_object, 
  group.by = "clust_SEU_d30_r0.5", 
  cols = clust_colors, 
  fov = Images(seurat_object), 
  combine = F)

names(plot_list) <- unique(seurat_object$orig.ident)

plot_list <- lapply(names(plot_list), function(name) {
  x <- plot_list[[name]]
  x <- x + 
    coord_flip() + 
    scale_x_reverse() + 
    ggtitle(name)
  })

names(plot_list) <- unique(seurat_object$orig.ident)

plot_combo <- patchwork::wrap_plots(plot_list) + plot_layout(guides = "collect")
plot_combo
```

### Feature Plot

```{r}
FeaturePlot(seurat_object, 
  features = c("Drd1", "Slc17a7"))
```

## Filtering the object

Once a user makes a choice of which reductions / parameters to keep for downstream analysis, we encourage to filter down the object.

Our [helper function](https://github.com/U-BDS/nf_xpatial/blob/main/assets/filter_xenium_obj.R) can be implemented in the following manner (please include the source code linked above in your environment). We provide a couple of different examples below as this function can be implemented to retain either multiple parameters or a single one:

```{r, echo = FALSE}
filter_xenium_obj <- function(
    input = NULL,
    assays = "Xenium",
    clustering_method = vector(),
    hmy_dims = vector(),
    hmy_res = vector(),
    bsky_dims = vector(),
    bsky_res = vector(),
    bsky_lambda = vector(),
    bsky_kgeom = vector() ) {

    metadata_cols <- colnames(input@meta.data)
    cluster_cols_keep <- metadata_cols[!grepl("^clust_", metadata_cols)]

    reductions_keep <- c()

    # Helper: build parameter patterns for regex matching
    add_param_to_keep <- function(prefix, param_list = vector() ) {

        suffix <- ifelse(length(param_list), param_list, ".*" )

        return(
          apply(expand.grid(prefix, suffix), 1, function(x) paste(x, collapse = ""))
        )
    }

    if ("Seurat" %in% clustering_method) {
      # Get the Cluster columns
      srt_clusts <- c("clust_HMY")

      srt_clusts <- add_param_to_keep(paste0(srt_clusts, "_d"), hmy_dims)
      srt_clusts <- add_param_to_keep(paste0(srt_clusts, "_r"), hmy_res)

      cluster_cols_keep <- c(cluster_cols_keep, srt_clusts)

      # Get the reductions
      reductions_keep <- c(
        reductions_keep,
        add_param_to_keep("Harmony_.*_d", hmy_dims)
      )
    }

    if ("BANKSY" %in% clustering_method) {
      # Get the Cluster columns
      bsky_clusts <- c("clust_BSKY_AGF1")

      bsky_clusts <- add_param_to_keep(paste0(bsky_clusts, "_L"), bsky_lambda)
      bsky_clusts <- add_param_to_keep(paste0(bsky_clusts, "_k"), bsky_kgeom)
      bsky_clusts <- add_param_to_keep(paste0(bsky_clusts, "_d"), bsky_dims)
      bsky_clusts <- add_param_to_keep(paste0(bsky_clusts, "_R"), bsky_res)

      cluster_cols_keep <- c(cluster_cols_keep, bsky_clusts)

      # Get the reductions
      reductions_keep <- c(
        reductions_keep,
        add_param_to_keep("BANKSY_.*_d", bsky_dims)
      )
    }

    filtered_seurat <- input

    # Build regex patterns
    reductions_keep_regex <- paste(reductions_keep, collapse = "|")
    cluster_keep_regex <- paste(cluster_cols_keep, collapse = "|")

    # Filter reductions
    all_reductions <- Reductions(input)
    reductions_to_remove <- all_reductions[!grepl(reductions_keep_regex, all_reductions)]

    for (reduction in reductions_to_remove) {
      filtered_seurat[[reduction]] <- NULL
    }

    # Filter metadata columns
    cols_to_remove <- metadata_cols[!grepl(cluster_keep_regex, metadata_cols)]

    for (col in cols_to_remove) {
      filtered_seurat[[col]] <- NULL
    }

    return(filtered_seurat)

  }
```

```{r}
# This will grab all the BANKSY cluster and reductions, filtering out the Seurat cluster results
only_banksy_obj <- filter_xenium_obj(
  input = seurat_object,
  clustering_method = c("BANKSY")
)
only_banksy_obj

# This will grab only the Seurat cluster and reductions for dimension 20 and resolution 0.4
single_seurat <- filter_xenium_obj(
  input = seurat_object,
  clustering_method = c("Seurat"),
  hmy_dims = c(20),
  hmy_res = c(0.4)
)
single_seurat

# This will grab only the Seurat cluster and reductions for dimension 20, 25,and 30 and resolution 0.4
mutliple_dims_seurat <- filter_xenium_obj(
  input = seurat_object,
  clustering_method = c("Seurat"),
  hmy_dims = c(20,25,30),
  hmy_res = c(0.4)
)
mutliple_dims_seurat
```